Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-03-29 paper/synthesis.tex
Done Section 7

When full computational power is needed, we can use the expressive
power of AGREE code contracts to write arbitrarily complex filter or
monitor operations, as we have already seen. Our formal model of AGREE
syntax and semantics is applied to extract code contracts and map them
to equivalent logic functions. In particular, given a code contract with
functionality ${\Eqs}\cdot G$, input ports $\mathbb{I}$, state
variables $\mathbb{S}$, and output ports $\mathbb{O}$, the semantics
given in Section \ref{sec:code-contracts} can be applied to create a
function
\[
 \konst{strmStep} : \mathbb{N}_0 \to \mathit{Env} \to \mathit{Env} = \sem{{\Eqs}\cdot G}
\]
Assume ${\Eqs}\cdot G$ is \emph{Pascalish}, \ie, sequentially
imperative. Then we can direcctly create a corresponding logic function
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}
\]
that takes input and state as parameters and returns an updated state
and output. Thus the environment $\mathit{Env}$ holding all previous
values of each variable can be replaced by \emph{parameters}, each of
which is able to hold only one data element. Assuming
$\mathit{Env}' = \konst{strmStep}\;t\; \mathit{Env}$, we can formally show that
\[
\vdash (\mathit{Env}' \vert_{\mathbb{S}},
 \mathit{Env}' \vert_{\mathbb{O}}) =
\konst{stateStep} (\mathit{Env}\vert_{\mathbb{I}(t)},
                   \mathit{Env}\vert_{\mathbb{S}(t)})
\]
In other words, projecting out the state and output components from
$\konst{strmStep}\;t\; \mathit{Env}$ equals the result of
applying \konst{stateStep} on the input and state components of
$\mathit{Env}$ at time $t$. This theorem connects the stream semantics
of the code contract to a pure logic function over the values
appearing in the stream. From this point we can use tools in the
CakeML eco-system to jump from the setting of higher order logic to
CakeML abstract syntax \cite{cakeml-translator}. Thus we have moved
from stream semantics to logic function to CakeML evaluation semantics
in a property-preserving way. Once in CakeML, further necessary
features can be added, namely imperative variables for storing state
between component invocations, also invocations of the (verified) foreign
function interface of CakeML. Finally, the CakeML compiler can be
applied to the program. It delivers an executable which can be shown
to preserve the original contract.

%% where the $\mathit{state}$ parameters to the function are just the
%% variables defined in ${\Eqs}$, obtained by projecting out the

%%  = \mathit{Env}\vert_{\mathbb{I}}$ $\mathit{input}
%%  = \mathit{Env}\vert_{\mathbb{I}}$ $\mathit{input}
%%  = \mathit{Env}\vert_{\mathbb{I}}$

@@

\begin{remark}[Partiality]

Partiality is an important consideration: steps 2 and 3 above can
fail; the data might not be parseable or the wellformedness
computation could be badly written and fail at runtime. In such cases,
the filter should recover and yield control without passing the input
onwards. In these cases, the filter is behaving as it should, but we
must also guard against situations in which a \emph{correctly
specified} filter fails at runtime. This kind of defect arises when
the filter \emph{ought} to accept a message, but lack of resources
results in the filter failing to do so. For example, the parse of a
message might need more space than has been allocated; another example
could be if the time slice provided by the scheduler is too short for
the wellformedness computation to finish. Thus resource bounds need to
be included in the correctness argument. Some preliminary work that
could support formal proofs of such bounds has been done for
CakeML \cite{cakeml-space-cost}.

\end{remark}

@@

Monitors are intended to track and analyze the externally visible
behavior of system components through time. Therefore, they tend to
require more extensive computational ability than filters. In
particular, our basic notion of a monitor is that it embodies a
relation over its input and output streams, and is able to access the
value of a stream at any earlier point in time, if necessary. Monitors
commonly use state to keep track of earlier values, unlike filters
which, for us, are typically stateless components. (However, there is
nothing in our approach that forbids stateful filters: they can be
realized by monitors.) A monitor specification is mapped by code
generation to a state transformation function of the following
abstract type:
%
\[
\konst{stepFn} : \mathit{input} \times \mathit{stateVars} \to \mathit{stateVars} \times \mathit{output}
\]
The \konst{stepFn} works as follows:
\begin{enumerate}

\item Each input is parsed into data of the type specified by the port
  type;

\item New values for the state variables are computed, in dependency
  order. The discussion above on initialization now comes into
  play. Suppose the variable declarations have the following form:
\[
\begin{array}{l}
  v_1 = i_1 \longrightarrow e_1 \\
  \cdots \\
  v_n = i_n \longrightarrow e_n \\
\end{array}
\]
In the generated code, for the first invocation of \konst{stepFn} only,
the initializations are executed in order:
\[
\begin{array}{l}
  v_1 = i_1; \\
  \cdots \\
  v_n = i_n; \\
\end{array}
\]
In all subsequent steps, the \emph{non-initialization} assignments are performed:
\[
\begin{array}{l}
  v_1 = e_1; \\
  \cdots \\
  v_n = e_n; \\
\end{array}
\]

@@

\newsavebox{\monFn}
\begin{lrbox}{\monFn}
\begin{lstlisting}[style=myML]
fun stepFn (response,request)
    (rsp,req,preIsPending,isPending,latency,policy,alert)
let val stateVars' =
      if !initStep then
      let val rsp = Option.isSome response
          val req = Option.isSome request
          val preIsPending = False
          val isPending = req andalso not rsp
          val latency = 0
          val policy = (if rsp then req else True)
          val alert = not(policy)
          val () = (initStep := False)
      in
        (rsp,req,preIsPending,isPending,latency,policy,alert)
      end
      else
      let ...
          val rsp = Option.isSome response
          val req = Option.isSome request
          val preIsPending = pre isPending
          val isPending = (req andalso not rsp) orelse
                          ((not rsp) andalso pre isPending)
          val latency = (if req then 0 else (pre latency + 1)
          val policy =
            ((if isPending then
                (latency < Defs.max_latency) else True)
              andalso
              (if rsp then
                 (req orelse preIsPending) else True))
          val alert = (Defs.is_latched andalso (pre alert))
                      orelse not(policy)
      in
        (rsp,req,preIsPending,isPending,latency,policy,alert)
      end
   val (rsp',_,_,_,_,_,alert') = stateVars'
   val alertPort = if alert' then Some () else None
   val outputPort =
     if alert' then None else
     if rsp' then Some response
     else None
in
   (stateVars', (alertPort,outputPort))
end
\end{lstlisting}
\end{lrbox}

@@

The system scheduler \emph{activates} components in some order. It is
an obligation on the system that the scheduler follows some sensible
partial order of component activation and allows each component
sufficient time for its computation.  Activating a monitor component
takes the form of the following pseudo-code, in which the monitor
evaluates the \konst{stepFn} on its current inputs and the current
values of the state variables, returning the new state and the output
values.
\[
\begin{array}{ll}
 \mathit{(i_1,\ldots)} & = \konst{readInputs}(); \\
 (v_1,\ldots) & = \konst{readState}() ; \\
 ({v_1}',\ldots), ({o_1}',\ldots) & = \konst{stepFn} ((i_1,\ldots),(v_1,\ldots)) ; \\
 \multicolumn{2}{l}{\konst{writeState}({v_1}',\ldots);} \\
 \multicolumn{2}{l}{\konst{writeOutputs}({o_1}'\ldots);} \\
\end{array}
\]

>>>>

When full computational power is needed, we can use the expressive
power of AGREE code contracts to write arbitrarily complex filter or
monitor operations, as we have already seen. Our formal model of AGREE
syntax and semantics is applied to extract code contracts and map them
to equivalent logic functions. In particular, given a code contract with
functionality ${\Eqs}\cdot G$, input ports $\mathbb{I}$, state
variables $\mathbb{S}$, and output ports $\mathbb{O}$, the semantics
given in Section \ref{sec:code-contracts} can be applied to create a
function
\[
 \konst{strmStep} : \mathbb{N}_0 \to \mathit{Env} \to \mathit{Env} = \sem{{\Eqs}\cdot G}
\]
Assume ${\Eqs}\cdot G$ is \emph{Pascalish}, \ie, sequentially
imperative. Then we can direcctly create a corresponding logic function
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}
\]
that takes input and state as parameters and returns an updated state
and output. Thus the environment $\mathit{Env}$ holding all previous
values of each variable can be replaced by \emph{parameters}, each of
which is able to hold only one data element. Assuming
$\mathit{Env}' = \konst{strmStep}\;t\; \mathit{Env}$, we can formally prove that
\[
\vdash (\mathit{Env}' \vert_{\mathbb{S}},
 \mathit{Env}' \vert_{\mathbb{O}}) =
\konst{stateStep} (\mathit{Env}\vert_{\mathbb{I}(t)},
                   \mathit{Env}\vert_{\mathbb{S}(t)})
\]
In other words, projecting out the state and output components from
$\konst{strmStep}\;t\; \mathit{Env}$ equals the result of
applying \konst{stateStep} on the input and state components of
$\mathit{Env}$ at time $t$. This theorem connects the stream semantics
of the code contract to a pure logic function over the values
appearing in the stream. From this point we can use tools in the
CakeML eco-system to jump from the setting of higher order logic to
CakeML abstract syntax \cite{cakeml-translator}. Thus we have moved
from stream semantics to logic function to CakeML evaluation semantics
in a property-preserving way. Once in CakeML, further necessary
features can be added, namely imperative variables for storing state
between component invocations, also invocations of the (verified) foreign
function interface of CakeML \cite{cakeml-monadic}. Finally, the CakeML compiler can be
applied to the program. It delivers an executable which can be shown
to preserve the original contract.

@@

@@

%% Monitors are intended to track and analyze the externally visible
%% behavior of system components through time. Therefore, they tend to
%% require more extensive computational ability than filters. In
%% particular, our basic notion of a monitor is that it embodies a
%% relation over its input and output streams, and is able to access the
%% value of a stream at any earlier point in time, if necessary. Monitors
%% commonly use state to keep track of earlier values, unlike filters
%% which, for us, are typically stateless components. (However, there is
%% nothing in our approach that forbids stateful filters: they can be
%% realized by monitors.)
As discussed above, a monitor specification is
mapped to a state transformation function
%
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}.
\]
which computes as follows:
\begin{enumerate}

\item Each input buffer is parsed into data of the type specified by the port
  type;

\item New values for the state variables are computed, in dependency
  order. At this point variables defined by \emph{followd-by} are
  dealt with. Suppose the variable definitions have the following
  form:
\[
\begin{array}{l}
  v_1 = i_1 \longrightarrow e_1 \\
  \cdots \\
  v_n = i_n \longrightarrow e_n \\
\end{array}
\]
In the generated function, for the first invocation of \konst{stateStep} only,
the initializations are executed in order:
\[
  v_1 = i_1; \cdots ;  v_n = i_n;
\]
In all later invocations, the \emph{non-initialization} assignments are performed, also in order:
\[
  v_1 = e_1; \cdots ; v_n = e_n; \\
\]

@@

\newsavebox{\monFn}
\begin{lrbox}{\monFn}
\begin{lstlisting}[style=myML]
fun stateStep (response,request)
    (rsp,req,preIsPending,isPending,latency,policy,alert)
let val stateVars' =
      if !initStep then
      let val rsp = Option.isSome response
          val req = Option.isSome request
          val preIsPending = False
          val isPending = req andalso not rsp
          val latency = 0
          val policy = (if rsp then req else True)
          val alert = not(policy)
          val () = (initStep := False)
      in
        (rsp,req,preIsPending,isPending,latency,policy,alert)
      end
      else
      let ...
          val rsp = Option.isSome response
          val req = Option.isSome request
          val preIsPending = pre isPending
          val isPending = (req andalso not rsp) orelse
                          ((not rsp) andalso pre isPending)
          val latency = (if req then 0 else (pre latency + 1)
          val policy =
            ((if isPending then
                (latency < Defs.max_latency) else True)
              andalso
              (if rsp then
                 (req orelse preIsPending) else True))
          val alert = (Defs.is_latched andalso (pre alert))
                      orelse not(policy)
      in
        (rsp,req,preIsPending,isPending,latency,policy,alert)
      end
   val (rsp',_,_,_,_,_,alert') = stateVars'
   val alertPort = if alert' then Some () else None
   val outputPort =
     if alert' then None else
     if rsp' then Some response
     else None
in
   (stateVars', (alertPort,outputPort))
end
\end{lstlisting}
\end{lrbox}

@@

The system scheduler \emph{activates} components in some order. It is
an obligation on the system that the scheduler follows some sensible
partial order of component activation and allows each component
sufficient time for its computation.  Activating a monitor component
takes the form of the following pseudo-code, in which the monitor
evaluates the \konst{stateStep} on its current inputs and the current
values of the state variables, returning the new state and the output
values.
\[
\begin{array}{ll}
 \mathit{(i_1,\ldots)} & = \konst{readInputs}(); \\
 (v_1,\ldots) & = \konst{readState}() ; \\
 ({v_1}',\ldots), ({o_1}',\ldots) & = \konst{stateStep} ((i_1,\ldots),(v_1,\ldots)) ; \\
 \multicolumn{2}{l}{\konst{writeState}({v_1}',\ldots);} \\
 \multicolumn{2}{l}{\konst{writeOutputs}({o_1}'\ldots);} \\
\end{array}
\]

\paragraph{Partiality.} Partiality is an important consideration for both filters and
monitors: the input data might not be parseable or the wellformedness
computation of a filter could be badly written and fail at runtime. In
such cases, the filter should recover and yield control without
passing the input onwards. For these failure cases, the filter is
behaving as it should, but we must also guard against situations in
which a \emph{correctly specified} filter or monitor fails at
runtime. This kind of defect arises when a filter, for
example, \emph{ought} to accept a message, but lack of resources
results in the filter failing to do so. For example, the parse of a
message might need more space than has been allocated; another example
could be if the time slice provided by the scheduler is too short for
the wellformedness computation to finish. Thus resource bounds need to
be included in the correctness argument. Some preliminary work that
could support formal proofs of such bounds has been done for
CakeML \cite{cakeml-space-cost}.
